Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-07-07 paper/ijm.tex
more rewrite, till 6.2

\def\tenpc{$^{\ast}$}
\def\fivepc{$^{\ast\ast}$}
\def\onepc{$^{\ast\ast\ast}$}
\newcommand{\legend}{\normalsize{Significance levels:\hspace{1em} \tenpc : 10\% \hspace{1em} \fivepc : 5\% \hspace{1em} \onepc : 1\% \normalsize}}
\def\sep{0.5em}
\def\legendTwo{\multicolumn{3}{l}{\footnotesize{Significance levels
			:\hspace{1em} $\dag$ : 10\% \hspace{1em}
			$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
%SKB: Added for documenting the tables and figures
\newcommand{\modelTwo}{age, age\textsuperscript{2}, education, being married, number of children, inter-ethnic household}
\newcommand{\modelThree}{\modelTwo, industry, occupation ,ownership type of company, number of workers in company, working in the public sector, experience in company}
\newcommand{\modelThreeAdd}{industry, occupation, ownership type of company, number of workers in company, working in the public sector, experience in company}	
\newcommand{\restrictions}{The sample period is from year 2000 to year 2012. Sample is limited to persons
	25-55 year old.}
\newcommand{\agerestrictions}{Sample is limited to persons
	25-55 year old.}
%Modified section numbering	
\renewcommand{\thesection}{\arabic{section}.}
\renewcommand{\thesubsection}{\thesection\arabic{subsection}.}

@@

We analyse the relationship between labour market outcomes and language
skills by ordinary multivariate regression models.
Several related studies use instrumental variables to correct for
omitted variable bias and measurement errors \citep{Chiswick1995, bleakley+chin2004}.
We refrain from doing that. Our decision is mainly data driven,
there are no suitable instruments in our data,\footnote{We run an IV
	regression using age--at--immigration to instrument Estonian skills
	for those Russians who were in the country at 1990.  The results are
	not statistically significant at any conventional level.} but it also facilitates
comparison with other studies.  For instance,
\citet{azam+2013EDandCC} and \citet{paolo+tansel2015JofDevStud} only
report OLS results, and, even more, find that the available proxy
variables have little influence on the final results.

@@

We estimate the basic model in the following form:
\begin{equation}
	\label{eq:specification}
	\begin{split}
		y_{it} ^{g} = &\: \alpha^{g}{}' \vec{X}_{it}^{g} + \vec{\beta}^{g}{}' \cdot \vec{L}_{it}^{g} +
		\eta_{i}^{g} +
		\upsilon_{t}^{g} + \epsilon_{it}^{g},
	\end{split}
\end{equation}
where $y_{it}^{g}$ is the outcome (either unemployment indicator
or logarithm of net monthly wage)
of individual $i$ in year $t$ from ethnic group $g$ (either Estonian
or Russian); $\vec{X}$ are the other individual
characteristics; $\vec{L}$ is the vector of
language skill descriptors, $\eta_{i}^{g}$ is the individual-specific effect, $\upsilon_{t}^{g}$ are the year dummies and $\epsilon_{it}^{g}$
is the idiosyncratic error term. 
\fxerror{ot: what's eta?  do you do random effects?}
As we use panel data, we cluster the
standard errors on individuals.  When analysing the time trend, we
replace the main language effects in~\eqref{eq:specification} by the
corresponding interactions
\fxwarning{ot: ensure this is what we do :-)}
\begin{align}
	\label{eq:time_interaction}
	\beta_{RUS,t} \cdot RUS_{it} \cdot \upsilon_{t} +
	\beta_{EST,t} \cdot EST_{it} \cdot \upsilon_{t} +
	\beta_{ENG,t} \cdot ENG_{it} \cdot \upsilon_{t}.
\end{align}

@@

The central results for unemployment regressions are given in
\cref{tab:unemployment_estimation_by_sex_and_ethnic_same_sample}.
\footnote{This table is restricted to have the same number of observations across all models requested by reviewer. The unrestricted table is in the appendix \cref{tab:unemployment_estimation_by_sex_and_ethnic}.}

The
table clearly indicates that Estonian language skills are associated
with approximately 5 percentage points lower unemployment for both
Russian men and women. We also see that the effect is not radically
influenced by the individual characteristics--for men it is
virtually unchanged while for women the estimate falls by 30\%.

@@

Neither Russian nor English show a comparably strong effect.
Knowledge of Russian is not associated with less unemployment for
Estonian men, and is even related to a slightly higher unemployment
for Estonian women. The latter outcome may be related to omitted
variable bias, Russian knowledge was widespread before 1990 and may
not have been closely associated with any favourable characteristics,
but during the period was increasingly more associated with age.
English skills are related to less unemployment both for Estonian men
and women, but the effect is small.
For Russians, however, Estonian language skills matter. Those who are
fluent in that language have approximately 5 percentage point less
unemployment, the effect changes only a little when controlling for
individual characteristics.  In contrast, the effect of English
fluency is close to zero and not statistically significant.  As in
case of Estonian, individual characteristics change the estimates only
a little.

In summary, these results suggest that the Estonian language is
important for employment prospects. The role of English is
small, and that of Russian is almost non-existent despite a largely
bilingual service sector in the country.

\subsection{Wage Regressions}
%SKB: Adding occupation to Model 3 reduces the effects drastically. Beforehand I somehow forgot to add it to the estimation. Now occupation is included as it should have been in the first place.
\label{subsec:basic_model_wage}

\cref{tab:wage_estimation_by_sex_and_ethnic_same_sample} presents the main
results.\footnote{This table is restricted to have the same number of observations across all models requested by reviewer. The unrestricted table is in the appendix \cref{tab:wage_estimation_by_sex_and_ethnic}.}

The table
confirms that wage is positively associated with fluency in other
languages, in particular English. The only remarkable exception is
Estonian language for Russian men where we fail to find any positive
effect for any model. This qualitatively repeats the results of
\citet{Toomet2011}. However, for women we observe a sizeable
positive effect between 6 and 16 percentage points, suggesting that
\citet{Toomet2011} results only describe the male labour market.  The effect decreases somewhat
(from 0.167 to 0.114) when including the individual characteristics
into the model, suggesting that the premium
is partly caused by education and potentially also by unobserved
individual performance. Adding further observable workplace characteristics nearly halves the effects 
(when accounting not only for the industry but also the occupation) %SKB: Might remove this sentence
.

\begin{table}[htbp]

@@

We also see that knowledge of Russian language is associated with a
substantial income premium of 10\% for both Estonian men and women.
The effect approximately halves when introducing
individual controls, suggesting that it is partly related to
education and potentially also to ability. It
falls further when also controlling for the
workplace characteristics, indicating that Estonians' income is
better characterized by available workplace
variables.

@@

In conclusion, the wage regressions suggest that English is the main
language that matters for men.  The effect of Russian is minor and
that of Estonian non-existent.  For women, all three languages are
associated with higher wage but Russian to a smaller extent.
This is in a stark contrast to the unemployment regressions, we return to
this question in the discussion.

@@

Here we look at the yearly estimates from 1992 to 2012.
We report those for the model 2.\footnote{The results for other models
	look essentially the same. If we add more variables like in model 3 then the effect size reduces by about five to ten percentage points, but the overall picture remains the same.}

The effect on language fluency on unemployment is shown on Figure~\ref{fig:long-run_unemployment} and on wage on Figure~\ref{fig:long-run_wage}

\begin{figure}[hbtp]
	\centering
	\subfigure[Non-native languages for men]{
		\includegraphics[width=0.45\linewidth]{Figure1a.eps}

@@

The results reveal a few interesting facts.  First, the effects are
relatively stable over time, indicating that neither economic
environment nor the data quality have changed rapidly.  Estonian
fluency has been associated with less unemployment for both men and
women since around year 2000, in 1990s it had little effect.  Russian fluency,
in contrast, had never an impact on unemployment probability.

For Estonian men, fluency in English is associated with less unemployment through the whole
period of observation whereas fluency in Russian has no effect.
Only few of the results are statistically significant pointwise though.
For Estonian women, both languages are associated with less
unemployment in the 1990s but not later. The picture is different for
Russians. For men, we cannot see any substantial effect of fluency in English
whereas fluency in Estonian is clearly associated with less unemployment from
the late 1990s on. For women, the picture is mostly similar, just the
effect of Estonian language seems to be somewhat delayed compared to
that for men.

\begin{figure}[hbtp]
	\centering
	\subfigure[Non-native languages for men]{
		\includegraphics[width=0.45\linewidth]{Figure2a.eps}
		\label{fig:long-run_wage_estonian_russian_men}
	}
	\subfigure[Non-native languages for women]{
		\includegraphics[width=0.45\linewidth]{Figure2b.eps}
		\label{fig:long-run_wage_estonian_russian_women}
	}
	\subfigure[English for Men]{
		\includegraphics[width=0.45\linewidth]{Figure2c.eps}
		\label{fig:long-run_wage_english_men}
	}
	\subfigure[English for Women]{
		\includegraphics[width=0.45\linewidth]{Figure2d.eps}
		\label{fig:long-run_wage_english_women}
	}
	\caption{The effect of language fluency on wage, 1992--2012. \\ Control variables are \modelTwo. \agerestrictions}
	\label{fig:long-run_wage}
\end{figure}

The estimates for wage are rather stable as well. For Estonians, we
see a slight downward trend of the effect of fluency in Russian whereas
that of English becomes important in the late 1990s. For Russians, the
importance of English seems to be sliding across the years, probably
due to the increasing supply of these skills. For men, fluency in Estonian
has never been significantly associated with wage.
The results are remarkably stable over the whole period, stressing
the continuity of the peculiar labour market institutions,
potentially workplace segregation, through the
whole post-communist period. For women, however, we can see
a slight increasing trend, suggesting that there is either less
workplace segregation or other need for the official language for
the jobs commonly taken by women.

>>>>

\def\tenpc{$^{\ast}$}
\def\fivepc{$^{\ast\ast}$}
\def\onepc{$^{\ast\ast\ast}$}
\newcommand{\legend}{\normalsize{Significance levels:\hspace{1em} \tenpc : 10\% \hspace{1em} \fivepc : 5\% \hspace{1em} \onepc : 1\% \normalsize}}
\def\sep{0.5em}
\def\legendTwo{\multicolumn{3}{l}{\footnotesize{Significance levels
			:\hspace{1em} $\dag$ : 10\% \hspace{1em}
			$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
%SKB: Added for documenting the tables and figures
\newcommand{\modelTwo}{age, age\textsuperscript{2}, education, marriage, number of children, inter-ethnic household}
\newcommand{\modelThree}{\modelTwo, industry, occupation, ownership type of company, number of workers in company, working in the public sector, experience in company}
\newcommand{\modelThreeAdd}{industry, occupation, ownership type of company, number of workers in company, working in the public sector, experience in company}	
\newcommand{\restrictions}{The sample period is from year 2000 to year 2012. Sample is limited to persons
	25-55 year old.}
\newcommand{\agerestrictions}{25-55 year olds only.}
%Modified section numbering	
\renewcommand{\thesection}{\arabic{section}.}
\renewcommand{\thesubsection}{\thesection\arabic{subsection}.}

@@

We analyse the relationship between labour market outcomes and language
skills by ordinary multivariate regression models.
Several related studies use instrumental variables to correct for
omitted variable bias and measurement errors \citep{Chiswick1995, bleakley+chin2004}.
We refrain from doing that. Our decision is mainly data driven,
there are no suitable instruments in our data,\footnote{We run an IV
	regression using age at immigration to instrument Estonian skills
	for those Russians who were in the country at 1990.  The results are
	not statistically significant at any conventional level.} but it also facilitates
comparison with other studies.  For instance,
\citet{azam+2013EDandCC} and \citet{paolo+tansel2015JofDevStud} only
report OLS results, and, even more, find that the available proxy
variables have little influence on the final results.

@@

We estimate the basic model in the following form:
\begin{equation}
	\label{eq:specification}
	\begin{split}
		y_{it} ^{g} = &\: \alpha^{g}{}' \vec{X}_{it}^{g} + \vec{\beta}^{g}{}' \cdot \vec{L}_{it}^{g} +
		\eta_{i}^{g} +
		\upsilon_{t}^{g} + \epsilon_{it}^{g},
	\end{split}
\end{equation}
where $y_{it}^{g}$ is the outcome (either unemployment indicator
or logarithm of net monthly wage)
of individual $i$ in year $t$ from ethnic group $g$ (either Estonian
or Russian); $\vec{X}$ are the other individual
characteristics; $\vec{L}$ is the vector of
language skill descriptors, $\eta_{i}^{g}$ is the individual-specific effect, $\upsilon_{t}^{g}$ are the year dummies and $\epsilon_{it}^{g}$
is the idiosyncratic error term. 
\fxerror{ot: what's the eta term?  do you do random effects?}
As we use panel data, we cluster the
standard errors on individuals.  When analysing the time trend, we
replace the main language effects in~\eqref{eq:specification} by the
corresponding interactions
\fxwarning{ot: ensure this is what we do :-)}
\begin{align}
	\label{eq:time_interaction}
	\beta_{RUS,t} \cdot RUS_{it} \cdot \upsilon_{t} +
	\beta_{EST,t} \cdot EST_{it} \cdot \upsilon_{t} +
	\beta_{ENG,t} \cdot ENG_{it} \cdot \upsilon_{t}.
\end{align}

@@

The central results for unemployment regressions are given in
\cref{tab:unemployment_estimation_by_sex_and_ethnic_same_sample}.\footnote{We
  restrict both models to have the same number of observations as
  requested by a reviewer. The unrestricted table is in the appendix \cref{tab:unemployment_estimation_by_sex_and_ethnic}.}
The
table clearly indicates that Estonian language skills are associated
with approximately 5 percentage points lower unemployment for both
Russian men and women. We also see that the effect is not radically
influenced by the individual characteristics--for men it is
virtually unchanged while for women the estimate falls by 30\%, from
6.5 to 4.5 percentage points.

@@

Neither Russian nor English show a comparably strong effect.
Knowledge of Russian is not associated with less unemployment for
Estonian men, and is even related to a slightly higher unemployment
for Estonian women. The latter outcome may be related to omitted
variable bias, Russian knowledge was widespread in the Soviet era and may
not have been closely associated with any favourable characteristics,
but has been increasingly more associated with age  during the period of analysis.
English skills are related to less unemployment both for Estonian men
and women, but the effect is small.
For Russians, however, Estonian language skills matter. Those who are
fluent in that language have approximately 5 percentage point less
unemployment, the effect changes only a little when controlling for
individual characteristics.  In contrast, the effect of English
fluency is close to zero and not statistically significant.  As in
case of Estonian, individual characteristics change the estimates only
a little.

In summary, these results suggest that Estonian is the only language
that matters for the employment prospects.  Despite of widespread
bilingualism in the service sector, and increasing global importance of
English, these languages are not closely associated with unemployment
probability.

\subsection{Wage}
%SKB: Adding occupation to Model 3 reduces the effects drastically. Beforehand I somehow forgot to add it to the estimation. Now occupation is included as it should have been in the first place.
\label{subsec:basic_model_wage}

\cref{tab:wage_estimation_by_sex_and_ethnic_same_sample} presents the
central wage regression
results.\footnote{As above, we
  restrict both models to have the same number of observations as
  requested by a reviewer.  The unrestricted table is in the appendix \cref{tab:wage_estimation_by_sex_and_ethnic}.}
The table
confirms the descriptive evidence that wage is positively associated
with fluency of all analyzed languages with English showing the
largest effect.  The only remarkable exception is
Estonian language for Russian men where we fail to find any positive
effect for any of the specifications.  This qualitatively repeats the results of
\citet{Toomet2011}. However, for women we observe a sizeable
positive effect between 6 and 16 percentage points, suggesting that
\citet{Toomet2011} results are not valid for the female labour market.  The effect decreases somewhat
(from 0.167 to 0.114) when including the individual characteristics
into the model, suggesting that the premium
is partly caused by education and potentially also by unobserved
individual performance.  Adding observable workplace characteristics
lowers the effects further to 0.066, suggesting that part of the
language skill premium is collected through workplace mobility.

\begin{table}[tbp]

@@

We also see that knowledge of Russian language is associated with a
substantial income premium of 10\% for both Estonian men and women.
The effect falls rapidly when introducing
individual and workplace controls, suggesting that both ability bias
and workplace mobility play a role.

@@

In conclusion, the wage regressions suggest that English is the main
language that matters for men.  The effect of Russian is minor and
that of Estonian non-existent.  For women, all three languages are
associated with higher wage but Russian to a smaller extent.
This is in a stark contrast to the unemployment regressions and we return to
this question in the discussion below.

@@

As Estonian economy went through turbulent times and rapid changes
during the 1990s, additional evidence can be gained by a long-term
trends in the estimated effect size.  Here we look at the yearly estimates from 1992 to 2012.
We report the results for model 2, i.e. models that include year
dummies and individual controls, but not workplace
characteristics.\footnote{The results for models with workplace data (model
  3 for wage)
  look qualitatively similar.  More controls in the model reduces the effect size but does not qualitatively
  change it's temporal pattern.}

The effect of language fluency on unemployment is shown on
Figure~\ref{fig:long-run_unemployment}, and on wage on
Figure~\ref{fig:long-run_wage}.
\fxnote{ot: SKB, I think I asked if you can slightly offset the black
  and red lines and you could not?}
The results reveal a few interesting facts.  First, the effects are
relatively stable over time, indicating that neither economic
environment nor the data quality have changed rapidly.  Estonian
fluency has been associated with less unemployment for both Russian men and
women since around year 2000, in 1990s it had little effect.  Russian fluency,
in contrast, had never had an impact on Estonians' unemployment probability.

\begin{figure}[bt!]
	\centering
	\subfigure[Non-native languages for men]{
		\includegraphics[width=0.45\linewidth]{Figure1a.eps}

@@

For Estonian men, fluency in English is associated with less unemployment through the whole
period of observation whereas fluency in Russian has no effect.
Only few of the results are pointwise statistically significant though.
For Estonian women, both languages are associated with less
unemployment in the 1990s but not later. The picture is different for
Russians. For men, we cannot see any substantial effect of fluency in English
whereas fluency in Estonian is clearly associated with less unemployment from
the late 1990s on. For women, the picture is mostly similar, just the
effect of Estonian language seems to be somewhat delayed compared to
that for men.

\begin{figure}[tb!]
	\centering
	\subfigure[Non-native languages for men]{
		\includegraphics[width=0.45\linewidth]{Figure2a.eps}
		\label{fig:long-run_wage_estonian_russian_men}
	}
	\subfigure[Non-native languages for women]{
		\includegraphics[width=0.45\linewidth]{Figure2b.eps}
		\label{fig:long-run_wage_estonian_russian_women}
	}
	\subfigure[English for Men]{
		\includegraphics[width=0.45\linewidth]{Figure2c.eps}
		\label{fig:long-run_wage_english_men}
	}
	\subfigure[English for Women]{
		\includegraphics[width=0.45\linewidth]{Figure2d.eps}
		\label{fig:long-run_wage_english_women}
	}
	\caption{The effect of language fluency on wage, 1992--2012. \\ Control variables are \modelTwo. \agerestrictions}
	\label{fig:long-run_wage}
\end{figure}

The wage effects show a similar smooth long-term trend.  For Estonians, we
see a slowly fading effect of Russian fluency whereas
English becomes important in the late 1990s.  For Russian men, fluency in Estonian
has never been significantly associated with wage.
The results are remarkably stable over the whole period, indicating
that the peculiar labour market institutions have been persistent for
two decades.  However, we can see how Estonian language turned
important in late 1990s, suggesting that there is either less
workplace segregation or other need for the official language for
the jobs commonly taken by women.  The
importance of English seems to be falling over time for Russians, possibly
due to the increasing supply of these skills.
